\documentclass{article}
\pdfoutput=1

\input{header}
\input{defs}

\usepackage{appendix}
\usepackage{blkarray}

\title{The Partially Hidden Markov Model: Incorporating Sparse Labels in Biologging Studies with Weighted Likelihoods}

\author{
  \textbf{Evan Sidrow} \\
  Department of Statistics \\
  University of British Columbia\\
  Vancouver, Canada \\
  \texttt{evan.sidrow@stat.ubc.ca} \\
  %
  \and
  %
  \textbf{Nancy Heckman} \\
  Department of Statistics \\
  University of British Columbia \\
  Vancouver, Canada \\
  %
  \and
  %
  \textbf{Sarah M. E. Fortune} \\
  Department of Oceanography \\
  Dalhousie University \\
  Halifax, Canada \\
  %
  \and
  %
  \textbf{Andrew W. Trites} \\
  Department of Zoology \\
  Institute for the Oceans and Fisheries \\
  University of British Columbia \\
  Vancouver, Canada \\
  %
  \and
  %
  \textbf{Marie Auger-M\'eth\'e} \\
  Department of Statistics \\
  Institute for the Oceans and Fisheries \\
  University of British Columbia \\
  Vancouver, Canada \\
}

\begin{document}

\maketitle

\section{Introduction}
\label{sec:intro}
\input sections/intro.tex

\section{Background}
\label{sec:background}
\input sections/background.tex

\section{Weighted Likelihood for Semi-Supervised Learning in Hidden Markov Models}
\label{sec:model}
\input sections/model.tex

%\section{Simulation Study}
%\label{sec:sim}
%\input sections/sim_study.tex

\section{Case Studies}
\label{sec:case}
\input sections/case_study_1.tex
\input sections/case_study_2.tex

\section{Discussion}
\label{sec:discussion}
\input sections/discussion.tex

\section*{ACKNOWLEDGEMENTS}
The killer whale data were collected under University of British Columbia Animal Care Permit number A19-0053 and Fisheries and Oceans Canada Marine Mammal Scientific License for Whale Research number XMMS 6 2019.
This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO). 
We thank the University of British Columbia and the Four-Year Doctoral Fellowship program.
We acknowledge the Canadian Statistical Sciences Institute (CANSSI) for its support.
We appreciate the support of the Canada Research Chairs program.
We acknowledge support from the BC Knowledge Development Fund.
We thank Canada Foundation for Innovation for its support.
\newpage
\bibliography{references}
%

\begin{appendix}

\input sections/appendix.tex

\end{appendix}

\end{document}
